---
output: html_document
---

# Lab 9x: Programming in R - Control Structures and Functions

## Control structures
- if and else: testing a condition and acting on it
- for: execute a loop a fixed number of times
- while: execute a loop while a condition is true

Control structures are generally used when writing functions (not alone).

### Conditional statements with if-else

#### Example of an else-if structure
##### Original
```{r}

## Generate a uniform random number for x
x <- runif(1, 0, 10)

##  Create an if-else structure where y will equal 10 or 0 depending on the number that is randomly generated for x
if(x > 3) {
        y <- 10
} else {
        y <- 0
}

x
y

```

##### Modified
```{r}

## Generate a uniform random number for x
x <- runif(1, 0, 10)

##  As a proof a concept since the random number generator cannot generate a number > 10, the y will equal the second argument which is 0
if(x > 10) {
        y <- 10
} else {
        y <- 0
}

x
y

```



### for Loops
#### Example
##### Original
```{r}

for(i in 1:10) {
        print(i)
}

```

##### Modified
```{r}

## Modified version of above example
for(i in 1:10) {
        print(i^2)
}

```

#### Example
##### Original
```{r}

x <- c("a", "b", "c", "d")
for(i in 1:4) {
        ## Print out each element of 'x'
        print(x[i])  
}

```

##### Modified
```{r}

## Modified version of above example example
x <- c("a", "b", "c", "d")
for(i in 1:3) {
        ## Print out first three elements of 'x'
        print(x[i])  
}

```

#### Example
##### Original
```{r}

## Generate a sequence based on length of 'x'
for(i in seq_along(x)) {   
        print(x[i])
}

```

##### Modified
```{r}

## Generate a sequence based on length of 'x' - 1
for(i in seq_along(x)) {   
        print(x[i - 1])
}

```

#### Example
##### Original
```{r}

for(letter in x) {
        print(letter)
}
```

##### Modified
```{r}

for(briefe in x) {
        print(briefe)
}
```


#### Example
##### Original
```{r}

for(i in 1:4) print(x[i])

```

##### Modified
```{r}

for(i in 3:4) print(x[i])

```

### Nested for Loops
#### Example
##### Original
```{r}

x <- matrix(1:6, 2, 3)
x

for(i in seq_len(nrow(x))) {
        for(j in seq_len(ncol(x))) {
                print(x[i, j])
        }   
}

```

##### Modified
```{r}

x <- matrix(1:6, 2, 3)
x

## Print the first three digits in 'x' matrix since there are three columns ('j') in the matrix for a total of 6 lines
for(i in seq_len(nrow(x))) {
        for(j in seq_len(ncol(x))) {
                print(x[j])
        }   
}

```

```{r}

x <- matrix(1:6, 2, 3)
x

## Print the first two digits in 'x' matrix since there are three columns ('i') in the matrix for a total of 6 lines
for(i in seq_len(nrow(x))) {
        for(j in seq_len(ncol(x))) {
                print(x[i])
        }   
}

```

### while Loops
#### Example
##### Original
```{r}

count <- 0
while(count < 10) {
        print(count)
        count <- count + 1
}

```

##### Modified
```{r}

## List count + 2 but stop when count is no longer less than 20
count <- 5
while(count < 20) {
        print(count)
        count <- count + 2
}

```

#### Example
##### Original
```{r}

z <- 5
set.seed(1)
while(z >= 3 && z <= 10) {
        coin <- rbinom(1, 1, 0.5)
        
        if(coin == 1) {  ## random walk
                z <- z + 1
        } else {
                z <- z - 1
        } 
}
print(z)

```

##### Modified
```{r}

## when 'z' doesn't meet the conditions
z <- 1
set.seed(1)
while(z >= 3 && z <= 10) {
        coin <- rbinom(1, 1, 0.5)
        
        if(coin == 1) {  ## random walk
                z <- z + 1
        } else {
                z <- z - 1
        } 
}
print(z)

coin
```

```{r}

## since coin doesn't equal 1, 5 minus 6 equal -1
z <- 5
set.seed(1)
while(z >= 3 && z <= 10) {
        coin <- rbinom(1, 1, 0.5)
        
        if(coin == 1) {  ## random walk
                z <- z + 1
        } else {
                z <- z - 6
        } 
}
print(z)

```


## Functions
#### Example
##### Original
```{r}

f <- function() {
        cat("Hello, world!\n")
}

f()
```

##### Modified
```{r}

fun <- function() {
        cat("Hello, world!\n")
}

fun()
```

#### Example
##### Original
```{r}

f <- function(num) {
        for(i in seq_len(num)) {
                cat("Hello, world!/n")
                }
}

f(3)

```

##### Modified
```{r}

f <- function(num) {
        for(i in seq_len(num)) {
                cat("Hello, world!\n")
                }
}

f(6)

```

#### Example
##### Original
```{r}

f <- function(num) {
        hello <- "Hello, world!\n"
        for(i in seq_len(num)) {
                cat(hello)
        }
        chars <- nchar(hello) * num
        chars
}

meaningoflife <- f(3)

print(meaningoflife)

```

##### Modified
```{r}

f <- function(num) {
        hello <- "Hello!\n"
        for(i in seq_len(num)) {
                cat(hello)
        }
        chars <- nchar(hello) * num
        chars
}

meaningoflife <- f(2)

print(meaningoflife)

```

#### Example
##### Original
```{r}

f <- function(num = 1) {
        hello <- "Hello, world!\n"
        for(i in seq_len(num)) {
                cat(hello)
        }
        chars <- nchar(hello) * num
        chars
}

f() ## Use default value for 'num'

f(2)   ## Use user-specified value

```

##### Modified
```{r}

f <- function(num = 3) {
        hello <- "Hello, world!\n"
        for(i in seq_len(num)) {
                cat(hello)
        }
        chars <- nchar(hello) * num
        chars
}

f() ## Use default value for 'num'

f(2)   ## Use user-specified value

```

#### Example
##### Original
```{r}

f(num = 2)

```

##### Modified
```{r}

f(num = 4)

```
