---
output: html_document
---

# Lab 11x: Microbiome Analysis using phyloseq

#### We will complete the first 4 out of 6 tutorials below:
1. [Ordination](https://joey711.github.io/phyloseq/plot_ordination-examples.html)
2. [Alpha Diversity](https://jeffreyblanchard.github.io/EvoGeno/EvoGeno_Lab11x_phyloseq.html)
3. [Heatmaps](https://joey711.github.io/phyloseq/plot_heatmap-examples.html)
4. [Networks](http://joey711.github.io/phyloseq/plot_network-examples.html)
5. [Trees](https://joey711.github.io/phyloseq/plot_tree-examples.html)
6. [Bar Plots](https://joey711.github.io/phyloseq/plot_bar-examples.html)

## Ordination Tutorial

Load packages and data
```{r message=FALSE}

library("phyloseq"); packageVersion("phyloseq")
data(GlobalPatterns)
library("ggplot2"); packageVersion("ggplot2")
library("plyr"); packageVersion("plyr")

```

Change theme for all future plots to theme_bw
```{r}

theme_set(theme_bw())

```

There are tools in the phyloseq package for preprocessing data. You can see examples and details of preprocessing in this [dedicated preprocessing tutorial](http://joey711.github.io/phyloseq/preprocess.html). 

When preprocessing you must think hard about the preprocessing that you do and be able to explain why you did them. The steps you took during preprocessing should be clearly documented.

We will preprocess by first filtering low-occurrence, poorly-represented OTUs from the data. Below, the OTUs are removed and indicated not to appear more than 5 times in more than half the samples. 
```{r}

GP = GlobalPatterns
wh0 = genefilter_sample(GP, filterfun_sample(function(x) x > 5), A=0.5*nsamples(GP))
GP1 = prune_taxa(wh0, GP)

```

Next, we will transform the data to even the sampling depth.
```{r}

GP1 = transform_sample_counts(GP1, function(x) 1E6 * x/sum(x))

```

Next, we keep only the top five most abundant phyla.
```{r}

phylum.sum = tapply(taxa_sums(GP1), tax_table(GP1)[, "Phylum"], sum, na.rm=TRUE)
top5phyla = names(sort(phylum.sum, TRUE))[1:5]
GP1 = prune_taxa((tax_table(GP1)[, "Phylum"] %in% top5phyla), GP1)

```

The preprocessing still leaves us with 204 OTUs. A major problem here is that some are human-associated microbiomes and some are not. Next, we will define human-associated versus non-human.
```{r}

human = get_variable(GP1, "SampleType") %in% c("Feces", "Mock", "Skin", "Tongue")
sample_data(GP1)$human <- factor(human)

```

### Four Main Ordination Plots
The plot_ordination() function supports four basic representations of an ordination.

We'll start by plotting just the OTUs and coloring the points by Phylum.
```{r}

GP.ord <- ordinate(GP1, "NMDS", "bray")
p1 = plot_ordination(GP1, GP.ord, type="taxa", color="Phylum", title="taxa")
print(p1)

```

This is a complicated looking plot, which is not necessarily a good thing. We can use facet_wrap() function to simplify the plot.
```{r}

p1 + facet_wrap(~Phylum, 3)

```

Next, we'll plot only the samples and color the points by sample type while also modifying the shape according to whether they are human-associated or not.
```{r}

p2 = plot_ordination(GP1, GP.ord, type="samples", color="SampleType", shape="human") 
p2 + geom_polygon(aes(fill=SampleType)) + geom_point(size=5) + ggtitle("samples")

```

The plot_ordination() function also allows you to be able to plot both samples and OTUs on the same graph. 
```{r}

p3 = plot_ordination(GP1, GP.ord, type="biplot", color="SampleType", shape="Phylum", title="biplot")
# Some stuff to modify the automatic shape scale
GP1.shape.names = get_taxa_unique(GP1, "Phylum")
GP1.shape <- 15:(15 + length(GP1.shape.names) - 1)
names(GP1.shape) <- GP1.shape.names
GP1.shape["samples"] <- 16
p3 + scale_shape_manual(values=GP1.shape)

```

And, of course, we can simplify this graph (reduce occlusion) by using the argument type="split" instead of the facet_wrap approach.
```{r}

p4 = plot_ordination(GP1, GP.ord, type="split", color="Phylum", shape="human", label="SampleType", title="split") 
p4

```

### Supported Ordination Methods
Here we try different method parameter options to the plot_ordination() function. 

```{r}

dist = "bray"
ord_meths = c("DCA", "CCA", "RDA", "DPCoA", "NMDS", "MDS", "PCoA")
plist = llply(as.list(ord_meths), function(i, physeq, dist){
        ordi = ordinate(physeq, method=i, distance=dist)
        plot_ordination(physeq, ordi, "samples", color="SampleType")
}, GP1, dist)

```

Store the ordination methods plot results in a list.
```{r}

names(plist) <- ord_meths

```

Next, we extract the data from each of the individual plots and put the extracted data into one big dataframe so that we can include all plots in one graphic.
```{r}

pdataframe = ldply(plist, function(x){
    df = x$data[, 1:2]
    colnames(df) = c("Axis_1", "Axis_2")
    return(cbind(df, x$data))
})
names(pdataframe)[1] = "method"

```

Next, we can use our dataframe to make a standard faceted ggplot scatterplot.
```{r}

p = ggplot(pdataframe, aes(Axis_1, Axis_2, color=SampleType, shape=human, fill=SampleType))
p = p + geom_point(size=4) + geom_polygon()
p = p + facet_wrap(~method, scales="free")
p = p + scale_fill_brewer(type="qual", palette="Set1")
p = p + scale_colour_brewer(type="qual", palette="Set1")
p

```

To replot a larger version of an individual plot, we can print the plist from which the pdataframe was made. Here, we plot the second element in the list.
```{r}

plist[[2]] 

```

We can add some additional layers to improve the aesthetics of the plot.
```{r}

p = plist[[2]] + scale_colour_brewer(type="qual", palette="Set1")
p = p + scale_fill_brewer(type="qual", palette="Set1")
p = p + geom_point(size=5) + geom_polygon(aes(fill=SampleType))
p

```

### MDS ("PCoA") on Unifrac Distances
Next, we can use the ordinate() function to simultaneously perform weighted UniFrac and perform a Principal Component Analysis on the distance matrix. Then, we can pass the data and the ordination results to plot_ordination to create a default ggplot2 graphic.
```{r}

ordu = ordinate(GP1, "PCoA", "unifrac", weighted=TRUE)
plot_ordination(GP1, ordu, color="SampleType", shape="human")

```

Then, we add layers to improve the aesthetics of the plot.
```{r}

p = plot_ordination(GP1, ordu, color="SampleType", shape="human")
p = p + geom_point(size=7, alpha=0.75)
p = p + scale_colour_brewer(type="qual", palette="Set1")
p + ggtitle("MDS/PCoA on weighted-UniFrac distance, GlobalPatterns")

```

## Alpha Diversity Tutorial

Load packages and GlobalPatterns data
```{r}

library("phyloseq"); packageVersion("phyloseq")
data("GlobalPatterns")
library("ggplot2"); packageVersion("ggplot2")

```

Set the ggplot themes
```{r}

theme_set(theme_bw())
pal = "Set1"
scale_colour_discrete <-  function(palname=pal, ...){
  scale_colour_brewer(palette=palname, ...)
}
scale_fill_discrete <-  function(palname=pal, ...){
  scale_fill_brewer(palette=palname, ...)
}

```

### Prepare data
First, we prune the OTUs that are not present in any of the samples. Although it is tempting to trim noise right away, many richness estimates are modeled on singletons and doubletons in the abundance data. Therefore, it is important to leave the noise in the dataset if you want a meaningful estimate.
```{r}

GP <- prune_species(speciesSums(GlobalPatterns) > 0, GlobalPatterns)

```

### Plot Examples
Here, we plot the default graphic produced by plot_richness() function.
```{r}

plot_richness(GP)

```

Next, we specify a measures argument to the plot_richness() function. This will include the alpha-diversity measures that we're interested in (i.e. Chao1 and Shannon).
```{r}

plot_richness(GP, measures=c("Chao1", "Shannon"))

```

Next, we specify a sample variable on which to group/organize samples along the x axis. In this case, we use the SampleType variable.
```{r}

plot_richness(GP, x="SampleType", measures=c("Chao1", "Shannon"))

```

If we wanted to use an external variable in the plot that isn't in the GP dataset already (e.g. such as human-associated versus not human-associated), we first have to define this new variable as a factor.
```{r}

sampleData(GP)$human <- getVariable(GP, "SampleType") %in% c("Feces", "Mock", "Skin", "Tongue")

```

Then, we tell the plot_richness() function to map the new variable on the x axis and color the points according to the SampleType they belong to.
```{r}

plot_richness(GP, x="human", color="SampleType", measures=c("Chao1", "Shannon"))

```

If we wanted to merge samples from the SampleType environment and we would use the merge_samples() function to merge the samples.
```{r}

GPst = merge_samples(GP, "SampleType")
# repair variables that were damaged during merge (coerced to numeric)
sample_data(GPst)$SampleType <- factor(sample_names(GPst))
sample_data(GPst)$human <- as.logical(sample_data(GPst)$human)

```

Then, we plot the environment-merged version of the data, and add a geom_point() layer to change the size and transparency of the points.
```{r}

p = plot_richness(GPst, x="human", color="SampleType", measures=c("Chao1", "Shannon"))
p + geom_point(size=5, alpha=0.7)

```

### More details about ggplot2
To remove the original layer points, first we check which lists are present in p.
```{r}

p$layers

```

We can see that the first layer is the one specifying the original points. Next, we use negative indexing to remove that layer and add a new geom_point with larger point size.
```{r}

p$layers <- p$layers[-1]
p + geom_point(size=5, alpha=0.7)

```


## Heatmap Plots Tutorial

Load packages
```{r}

library("phyloseq"); packageVersion("phyloseq")
library("ggplot2"); packageVersion("ggplot2")

```

Set theme for ggplot
```{r}

theme_set(theme_bw())

```

Upload data and plot a heatmap of the top 300 most abundant bacteria taxa across all samples. Note that no prior preprocessing occurred, which is not recommended.
```{r}

data("GlobalPatterns")
gpt <- subset_taxa(GlobalPatterns, Kingdom=="Bacteria")
gpt <- prune_taxa(names(sort(taxa_sums(gpt),TRUE)[1:300]), gpt)
plot_heatmap(gpt, sample.label="SampleType")

```

### Adjusting plot aesthetics
Subset a smaller dataset based on the Archaeal phylum using the function subset_taxa().
```{r}

gpac <- subset_taxa(GlobalPatterns, Phylum=="Crenarchaeota")

```

Next, create a default heatmap using the plot_heatmap() function.
```{r}

plot_heatmap(gpac)

```

We can also re-label based on sample type.
```{r}

(p <- plot_heatmap(gpac, "NMDS", "bray", "SampleType", "Family"))

```

To change the axes titles, you can use the following code:
```{r}

p$scales$scales[[1]]$name <- "My X-Axis"
p$scales$scales[[2]]$name <- "My Y-Axis"
print(p)

```

In order to change the color scheme you can use the low and high arguments in the plot_heatmap() function and specify the colors.
```{r}

plot_heatmap(gpac, "NMDS", "bray", "SampleType", "Family", low="#000033", high="#CCFF66")

```

Here is a different variation using the colors dark-blue and red.
```{r}

plot_heatmap(gpac, "NMDS", "bray", "SampleType", "Family", low="#000033", high="#FF3300")

```

Here is another variation using a very dark-blue and a very-light blue color scheme.
```{r}

plot_heatmap(gpac, "NMDS", "bray", "SampleType", "Family", low="#000033", high="#66CCFF")

```

Here is a "dark on light" color scheme which can be done by indicating the color white for those with abundance values of 0.
```{r}

plot_heatmap(gpac, "NMDS", "bray", "SampleType", "Family", low="#66CCFF", high="#000033", na.value="white")

```

Here is a different variation where the near zero color is closer to a cream color while the other colors are a blue-grey gradient.
```{r}

plot_heatmap(gpac, "NMDS", "bray", "SampleType", "Family", low="#FFFFCC", high="#000033", na.value="white")

```

### Different Ordination Methods and Distances
Here we use the NMDS ordination on the jaccard distance.
```{r}

plot_heatmap(gpac, "NMDS", "jaccard")

```

Here we use the Detrended correspondence analysis (DCA)
```{r}

plot_heatmap(gpac, "DCA", "none", "SampleType", "Family")

```

Here we use the unconstrained redundancy analysis (aka Principle Component Analysis)
```{r}

plot_heatmap(gpac, "RDA", "none", "SampleType", "Family")

```

Here we use the PCoA/MDS ordination on the bray-curtis distance (the default distance).
```{r}

plot_heatmap(gpac, "PCoA", "bray", "SampleType", "Family")

```

Here we use the MDS/PCoA ordination on the Unweighted-UniFrac distance.
```{r}

plot_heatmap(gpac, "PCoA", "unifrac", "SampleType", "Family")

```

Here we use the weighted-UniFrac distance and MDS/PCoA ordination.
```{r}

plot_heatmap(gpac, "MDS", "unifrac", "SampleType", "Family", weighted=TRUE)

```

We can also create a heatmap using base-R graphics and the more common (but problematic) hierarchical clustering organization.
```{r}

heatmap(otu_table(gpac))

```

## Networks Tutorial

Load packages and enterotype data
```{r}

library(phyloseq); packageVersion("phyloseq")
packageVersion("ggplot2")
data(enterotype)

```

To ensure complete reducibility of the images in this tutorial, we can set the rand number generator seed.
```{r}

set.seed(711L)

```

Remove 9 samples for which no enterotype designation was assigned.
```{r}

enterotype = subset_samples(enterotype, !is.na(Enterotype))

```

Create a network plot using the function plot_net().
```{r}

plot_net(enterotype, maxdist = 0.4, point_label = "Sample_ID")

```

Here we plot some of the sample variables onto the network graphic as color and shape.
```{r}

plot_net(enterotype, maxdist = 0.3, color = "SeqTech", shape="Enterotype")

```

In the above network, the max distance was informed but arbitrarily chosen. We can change this value, which decreases the number of edges in the network.
```{r}

plot_net(enterotype, maxdist = 0.2, color = "SeqTech", shape="Enterotype")

```

Next, we create an igraph-based network based on the default distance method Jaccard and a maximum distance of 0.3 between connected nodes.
```{r}

ig <- make_network(enterotype, max.dist=0.3)

```

Then, we plot this network with the default settings.
```{r}

plot_network(ig, enterotype)

```

Next, we map some of the sample variables onto the graphic using color and shape.
```{r}

plot_network(ig, enterotype, color="SeqTech", shape="Enterotype", line_weight=0.4, label=NULL)

```

Again, the choice of maximum-distance and distance method was informed but arbitrarily chosen. Next, we lower the maximum distance and observe a decreasing number of edges in the network.
```{r}

ig <- make_network(enterotype, max.dist=0.2)
plot_network(ig, enterotype, color="SeqTech", shape="Enterotype", line_weight=0.4, label=NULL)

```

Lastly, we observe how the network changes when we use the Bray-Curtis method instead of the Jaccard method.
```{r}

ig <- make_network(enterotype, dist.fun="bray", max.dist=0.3)
plot_network(ig, enterotype, color="SeqTech", shape="Enterotype", line_weight=0.4, label=NULL)

```

